---
title: "Intro to dplyr"
output: html_document
date: "2022-12-04"
---

## why use dplyr?

-   great for data exploration and transformation
-   intuitive to write and read

## dplyr functionality

-   Five basic verbs: `filter`, `select`, `arrange`, `mutate`, `summarize` (plus `group-by`)

## create a local data frame

-   we use `as.tibble` to make a "local data frame" which is just a wrapper for a data frame that prints nicely

```{r}

# convert to local data frame
books_l <- tibble::as_tibble(books)

```

## filter: keep rows matching certain criteria

```{r}

# view all novels or memoirs read in 2019 

# base approach:
books_l[books_l$year==2019 & books_l$sub.genre=="Novel" | books_l$sub.genre=="Memoir",]

#dplyr approach:
filter(books_l, year==2019, sub.genre=="Novel" | sub.genre=="Memoir")
#even better:
filter(books_l, year==2019, sub.genre %in% c("Novel", "Memoir"))

```

## select: pick columns by name

```{r}

# select year, pages, and genre

#base approach
books_l[, c("year", "pages", "genre")]

#dplyr approach
select(books_l, year, pages, genre)

```

## using %\>% operator

-   the usual way of performing multiple operations in one line is to nest but %\>% is a lot neater and easier to read

```{r}

# nesting approach to select book, author, and sub.genre and filter for history
filter(select(books_l, book, author, sub.genre), sub.genre=="History")

#chaining method using %>%
books_l %>%
  select(book, author, sub.genre) %>%
  filter(sub.genre=="History")

```

# arrange: reorder rows

```{r}

# select book, author, and pages and sort by pages in ascending order
books_l %>% 
  select(book, book, author, pages) %>%
  arrange(pages)

#same but in descending order
books_l %>% 
  select(book, book, author, pages) %>%
  arrange(desc(pages))

```

# mutate: add new variables

-   create new variables that are functions of existing variables

```{r}

# create a new variable that assigns each book a length of short, medium, or long based on # of pages

books_l <- 
  books_l %>% 
  mutate(
    length = case_when(
      pages <=200 ~ "Short", 
      pages >200 & pages <500 ~ "Medium",
      pages >=500 ~ "Long"
    ))

```

# summarise: reduce variables to values

-   group data by some variable and then aggregate it with some function

```{r}

# find average book length by year

books_l %>%
  group_by(year) %>%
  summarise(ave_pages = mean(pages))

# calculate the ave # of pages of fiction and non-fiction books read each year

books_l %>%
  group_by(genre) %>%
  summarize(ave_pages = mean(pages))

# for each year, count total number of books read and sort in descending order

books_l %>%
  group_by(year) %>%
  summarise(book_count=n()) %>%
  arrange(desc(book_count))

# for each sub-genre, count total number of books and sort in descending order

books_l %>%
  group_by(sub.genre) %>%
  summarise(page_count=n()) %>%
  arrange(desc(page_count))

```
